Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-03 WSSSPE1_report/WSSSPE1_summary_report.tex
addressing Berriman comments

Because progress in scientific research is dependent on the quality and
accessibility of software at all levels, it is now critical to address many
challenges related to the development, deployment, and maintenance of reusable
software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists,
researchers, and students are able to learn and adopt software-related skills
and methodologies. Established researchers are already acquiring some of these
skills, and in particular a specialized class of software developers is
emerging in academic environments who are an integral and embedded part of
successful research teams. The WSSSPE workshop was intended to provide a forum
for discussion of the challenges, including both positions and experiences. The
short papers and debates have been archived to provide a basis for continued
discussion, and the workshop has fed into the collaborative writing of this
document. An estimate of 90 to 150 participants were present
throughout the day. Additional papers based on extended versions of workshop
submissions are expected. The level of interest in the workshop has led the
organizers, working with some of the submitters and attendees, to plan two
additional workshops (at the 2014 SciPy and SC14 conferences) and to turn the
workshop website into a community website that can be used as a focus for
further discussion and progress. Additionally, a minisymposium that aims to
further explore some of the key issues raised in WSSSPE is co-organized by a
WSSSPE1 participant at the 2014 SIAM Annual Meeting on ``Reliable
Computational Science.''

@@

@@

\begin{quote}
In practice, scientific software activities are part of an ecosystem
where key roles are held by developers, users, and funders. All three
groups supply resources to the ecosystem, as well as requirements that
bound it. Roughly following the example of NSF's Vision and Strategy
for Software~\cite{NSF_software_vision},
the ecosystem may be viewed as having challenges related to:

\begin{itemize}[leftmargin=0.2in]
\item the development process that leads to new (versions of) software
\begin{itemize}[leftmargin=0.2in]
\item how fundamental research in computer science or
  science/engineering domains is turned into reusable software
\item software created as a by-product of research
\item impact of computer science research on the development of
    scientific software and vice versa
\end{itemize}
\item the support and maintenance of existing software, including
  software engineering
\begin{itemize}[leftmargin=0.2in]
\item governance, business, and sustainability models
\item the role of community software repositories, their operation and
  sustainability
\end{itemize}
\item the role of open source communities or industry
\item use of the software
\begin{itemize}[leftmargin=0.2in]
\item growing communities
\item reproducibility, transparency needs that may be unique to science
\end{itemize}
\item policy issues, such as
\begin{itemize}[leftmargin=0.2in]
\item measuring usage and impact
\item software credit, attribution, incentive, and reward
\item career paths for developers and institutional roles
\item issues related to multiple organizations and multiple countries,
  such as intellectual property, licensing, etc.
\item mechanisms and venues for publishing software, and the role of
  publishers
\end{itemize}
\item education and training
\end{itemize}

@@

\begin{quote}
We invite short (4-page) position/experience reports that will be used
to organize panel and discussion sessions. These papers will be
archived by a third-party service, and provided DOIs. We encourage
submitters to license their papers under a Creative Commons license
that encourages sharing and remixing, as we will combine ideas (with
attribution) into the outcomes of the workshop.  An interactive site
will be created to link these papers and the workshop discussion, with
options for later comments and contributions. Contributions will be
peer-reviewed for relevance and originality before the links are added
to the workshop site; contributions will also be used to determine
discussion topics and panelists. We will also plan one or more papers
to be collaboratively developed by the contributors, based on the
panels and discussions.
\end{quote}

@@

In this subtopic, two papers included discussion of the definition of
sustainability.  Venters et al.~\cite{Venters_WSSSPE} mentioned that this is a
rather ambiguous concept, and that the lack of an accepted definition gets in
the way of integrating the concept into software engineering. They suggest that
sustainability is a non-functional requirement, and that the quality of
software architectures determines sustainability.  They then propose that
sustainability is a measure of a set of central quality attributes:
extensibility, interoperability, maintainability, portability, reusability, and
scalability. Finally, they develop an architecture evaluation framework based
on scenarios that help to illuminate how to measure quality or sustainability
at the architectural level.

@@

In conclusion, we find that there are distinctions in the characteristics and
the kinds of challenges faced between generic utility and specific scientific
domain software projects. In terms of career paths of original developers, it
appears that the career paths for utility software developers are broader
compared to the developers of scientific software.  However, it appears that in
the case of scientific software, core competencies of original developers tend
to be domain science who happen to be the coders of their software. As projects
as a whole, the small-scale, specific software have a lower risk and cost in
terms of sustaining as compared to utility software.  Often the sustainability
of such projects is achieved by the fact that the core developer or team
heavily utilizes the software for their science, eg.  \emph{R/qtl, PETSc}.
Furthermore, the development of scientific software requires more scientific
skills compared with those of utility software. This diminishes the need of
specialist programmers and practices in many cases where the bulk of
development is carried out by a domain scientist. With these inferences, we
believe that separate guidelines and/or principles for sustainability be
defined and practiced for each category of scientific software and they be
given separate treatment.

@@

>>>>

Because progress in scientific research is dependent on the quality of and
accessibility to software at all levels, it is now critical to address many
challenges related to the development, deployment, and maintenance of reusable
software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists,
researchers, and students are able to learn and adopt software-related skills
and methodologies. Established researchers are already acquiring some of these
skills, and in particular a specialized class of software developers is
emerging in academic environments who are an integral and embedded part of
successful research teams. The WSSSPE workshop was intended to provide a forum
for discussion of the challenges, including both positions and experiences. The
short papers and debates have been archived to provide a basis for continued
discussion, and the workshop has fed into the collaborative writing of this
document. An estimate of 90 to 150 participants were present
throughout the day. Additional papers based on extended versions of workshop
submissions are expected. The level of interest in the workshop has led the
organizers, working with some of the submitters and attendees, to plan two
additional workshops (at the 2014 SciPy and SC14 conferences) and to turn the
workshop website into a community website that can be used as a focus for
further discussion and progress. Additionally, a minisymposium that aims to
further explore some of the key issues raised in WSSSPE is co-organized by a
WSSSPE1 participant at the 2014 Society for Industrial and Applied Mathematics
(SIAM) Annual Meeting on ``Reliable
Computational Science.''

@@

The goal of this paper is to summarize the contents of the workshop papers
and discussion, and identify issues that future WSSSPE events will address.

@@

\begin{quote}
``In practice, scientific software activities are part of an ecosystem
where key roles are held by developers, users, and funders. All three
groups supply resources to the ecosystem, as well as requirements that
bound it. Roughly following the example of NSF's Vision and Strategy
for Software~\cite{NSF_software_vision},
the ecosystem may be viewed as having challenges related to:

\begin{itemize}[leftmargin=0.2in]
\item the development process that leads to new (versions of) software
\begin{itemize}[leftmargin=0.2in]
\item how fundamental research in computer science or
  science/engineering domains is turned into reusable software
\item software created as a by-product of research
\item impact of computer science research on the development of
    scientific software and vice versa
\end{itemize}
\item the support and maintenance of existing software, including
  software engineering
\begin{itemize}[leftmargin=0.2in]
\item governance, business, and sustainability models
\item the role of community software repositories, their operation and
  sustainability
\end{itemize}
\item the role of open source communities or industry
\item use of the software
\begin{itemize}[leftmargin=0.2in]
\item growing communities
\item reproducibility, transparency needs that may be unique to science
\end{itemize}
\item policy issues, such as
\begin{itemize}[leftmargin=0.2in]
\item measuring usage and impact
\item software credit, attribution, incentive, and reward
\item career paths for developers and institutional roles
\item issues related to multiple organizations and multiple countries,
  such as intellectual property, licensing, etc.
\item mechanisms and venues for publishing software, and the role of
  publishers
\end{itemize}
\item education and training''
\end{itemize}

@@

\begin{quote}
``We invite short (4-page) position/experience reports that will be used
to organize panel and discussion sessions. These papers will be
archived by a third-party service, and provided DOIs [Digital Object Identifiers].
We encourage
submitters to license their papers under a Creative Commons license
that encourages sharing and remixing, as we will combine ideas (with
attribution) into the outcomes of the workshop.  An interactive site
will be created to link these papers and the workshop discussion, with
options for later comments and contributions. Contributions will be
peer-reviewed for relevance and originality before the links are added
to the workshop site; contributions will also be used to determine
discussion topics and panelists. We will also plan one or more papers
to be collaboratively developed by the contributors, based on the
panels and discussions.''
\end{quote}

@@

In this subtopic, two papers included discussion of the definition of
sustainability.  Venters et al.~\cite{Venters_WSSSPE} mentioned that this is a
rather ambiguous concept, and that the lack of an accepted definition gets in
the way of integrating the concept into software engineering. They suggest that
sustainability is a non-functional requirement, and that the quality of
software architectures determines sustainability. They then propose that
sustainability is a consequence of a set of central quality attributes:
extensibility, interoperability, maintainability, portability, reusability, and
scalability. Finally, they develop an architecture evaluation framework based
on scenarios that help to illuminate how to measure quality or sustainability
at the architectural level.

@@

In conclusion, we find that there are distinctions in the characteristics and
the kinds of challenges faced between generic utility and specific scientific
domain software projects. In terms of career paths of original developers, it
appears that the career paths for utility software developers are broader
compared to the developers of scientific software.  However, it appears that in
the case of scientific software, core competencies of original developers tend
to be domain science who happen to be the coders of their software. As projects
as a whole, the small-scale, specific software have a lower risk and cost in
terms of sustaining as compared to utility software.  Often the sustainability
of such projects is achieved by the fact that the core developer or team
heavily utilizes the software for their science, e.g.,  \emph{R/qtl, PETSc}.
Furthermore, the development of scientific software requires more scientific
skills compared with those of utility software. This diminishes the need of
specialist programmers and practices in many cases where the bulk of
development is carried out by a domain scientist. With these inferences, we
believe that separate guidelines and/or principles for sustainability be
defined and practiced for each category of scientific software and they be
given separate treatment.

@@

\subsection{Top Issues}
